---
title: "R Tutorial at the WZB"
subtitle: "6 - Recap Quiz"
author: "Markus Konrad"
date: "November 29, 2018"
output:
  ioslides_presentation:
    logo: _all/wzbfirmierungquerschwarzmitschrift.png
    css: _all/custom.css
---

```{r, include=FALSE}
knitr::opts_chunk$set(R.options = list(max.print = 60))
```



# Recap quiz

## Recap quiz

A small R quiz to check yourself.

## Data types

You have survey data where participants express their consent on something with a five level score from *"I totally agree"* (5) to *"I totally disagree"* (1). Which data type would you choose to store this data?

- A: Character string vector
- B: Factor with 5 levels
- C: Numeric (integer) vector with values from 1 to 5
- D: Ordered factor with 5 *ordered* levels from 1 to 5

<small>Multiple options may apply.</small>

## Data types

You have survey data where participants express their consent on something with a five level score from *"I totally agree"* (5) to *"I totally disagree"* (1). Which data type would you choose to store this data?

- A: Character string vector
- B: Factor with 5 levels
- **C: Numeric (integer) vector with values from 1 to 5✓**
- **D: Ordered factor with 5 *ordered* levels from 1 to 5✓**


## What could be the problem...

What could be the problem, if you get an error message like this:

`Error in ...() : could not find function "..."`

- A: I made a typo in the function call.
- B: Something's wrong with my R installation. I have to re-install R.
- C: There's simply no such function and I either have to define one myself or find and load a package that contains it.
- D: I forgot to load a package with `library()`.

<small>Multiple options may apply.</small>

## What could be the problem...

What could be the problem, if you get an error message like this:

`Error in ...() : could not find function "..."`

- **A: I made a typo in the function call. ✓**
- B: Something's wrong with my R installation. I have to re-install R.
- **C: There's simply no such function and I either have to define one myself or find and load a package that contains it. ✓**
- **D: I forgot to load a package with `library()`. ✓**

## You have the following data...

You have the following data stored in an object `smoker`:

```{r, echo=FALSE}
(smoker <- c(TRUE, NA, NA, TRUE, FALSE, FALSE, FALSE, TRUE, NA))
```

What's the correct way of finding out the **portion of `NA`s in percent?**

- A: `mean(is.na(smoker)) * 100`
- B: `mean(smoker * 100, na.rm = TRUE)`
- C: `sum(!is.na(smoker)) / length(smoker) * 100`
- D: `sum(is.na(smoker)) / length(smoker) * 100`

<small>Multiple options may apply.</small>

## You have the following data...

You have the following data stored in an object `smoker`:

```{r, echo=FALSE}
(smoker <- c(TRUE, NA, NA, TRUE, FALSE, FALSE, FALSE, TRUE, NA))
```

What's the correct way of finding out the **portion of `NA`s in percent?**

- **A: `mean(is.na(smoker)) * 100` ✓**
- B: `mean(smoker * 100, na.rm = TRUE)`
- C: `sum(!is.na(smoker)) / length(smoker) * 100`
- **D: `sum(is.na(smoker)) / length(smoker) * 100` ✓**

## Logic {.smaller}

Assume you have the following data:

```{r}
age <- c(20, 35, 19, 51, 20)
smoker <- c(TRUE, TRUE, FALSE, TRUE, FALSE)
country <- factor(c('USA', 'GB', 'GB', 'DE', 'USA'))
```

<br>
What's the correct logical expression for "all from the US or any non-smokers younger than 30"?

- A: `country == 'USA' & (!smoker | age < 30)`
- B: `(!smoker & age < 30) | country == 'USA'`
- C: `country == 'USA' | (!smoker & age < 30)`
- D: `country == 'USA' & (country != 'USA' & smoker & age < 30)`

<small>Multiple options may apply.</small>

## Logic {.smaller}

Assume you have the following data:

```{r}
age <- c(20, 35, 19, 51, 20)
smoker <- c(TRUE, TRUE, FALSE, TRUE, FALSE)
country <- factor(c('USA', 'GB', 'GB', 'DE', 'USA'))
```

<br>
What's the correct logical expression for "all from the US or any non-smokers younger than 30"?

- A: `country == 'USA' & (!smoker | age < 30)`
- **B: `(!smoker & age < 30) | country == 'USA'` ✓**
- **C: `country == 'USA' | (!smoker & age < 30)` ✓**
- D: `country == 'USA' & (country != 'USA' & smoker & age < 30)`

## What's wrong here?

Find three problems within this code:

```{r, eval=FALSE}
x <- c(4, 2, -2, 1 -3, 4, 1)
2nd_object <- sum(x[(x >= 2)]
```

## What's wrong here?

Find three problems within this code:

```{r, eval=FALSE}
x <- c(4, 2, -2, 1 -3, 4, 1)
2nd_object <- sum(x[(x >= 2)]
```

<br>

- a comma is probably missing in the `x` vector <br> c(4, 2, -2, 1**,** -3, 4, 1)
- you can't name an object `2nd_object` (objects are not allowed to start with digits)
- the last closing paranthesis for `sum()` is missing

## Transforming data frames {.smaller}

Consider the `airquality` data set:

```{r}
head(airquality, 3)
```

What will be the output of the following statement:

```{r, eval=FALSE}
arrange(select(airquality, Ozone, Wind), desc(Wind))
```
<br>

- A: We'll get an error message.
- B: A subset of `airquality` containing only the variables `Ozone` and `Wind` and additionally a description of the `Wind` variable.
- C: A subset of `airquality` containing only the variables `Ozone` and `Wind`, ordered by `Wind` in descending order.
- D: A subset of `airquality` where we select only non-NA observations from `Ozone` and `Wind`, additionally ordered by `Wind`.


## Transforming data frames {.smaller}

Consider the `airquality` data set:

```{r}
head(airquality, 3)
```

What will be the output of the following statement:

```{r, eval=FALSE}
arrange(select(airquality, Ozone, Wind), desc(Wind))
```
<br>

- A: We'll get an error message.
- B: A subset of `airquality` containing only the variables `Ozone` and `Wind` and additionally a description of the `Wind` variable.
- **C: A subset of `airquality` containing only the variables `Ozone` and `Wind`, ordered by `Wind` in descending order. ✓**
- D: A subset of `airquality` where we select only non-NA observations from `Ozone` and `Wind`, additionally ordered by `Wind`.


## Job application data {.smaller}

You have the following job application data in a data frame named `appl`:

```{r, echo=FALSE}
(appl <- data.frame(application_id = 1:7,
                    gender = factor(c('f', 'f', 'm', NA, 'm', 'x', 'f')),
                    success = c(TRUE, FALSE, TRUE, TRUE, TRUE, NA, FALSE)))
```

What would be your steps to find the **ratio of successful applications of women** (denoted as `f` in `gender` variable)? Don't write code, describe the steps and the functions you would use. Describe how you handle the occurrence of NAs.

## Job application data {.smaller}

You have the following job application data in a data frame named `appl`:

```{r, echo=FALSE}
(appl <- data.frame(application_id = 1:7,
                    gender = factor(c('f', 'f', 'm', NA, 'm', 'x', 'f')),
                    success = c(TRUE, FALSE, TRUE, TRUE, TRUE, NA, FALSE)))
```

What would be your steps to find the **ratio of successful applications of women** (denoted as `f` in `gender` variable)?

1. Get the number of female applicants by using `sum()` on a logical expression `appl$gender == 'f'` and setting `sum()` to ignore NAs (use `na.rm = TRUE` parameter).
2. Get the number of *successful* female applicants by subsetting the `success` variable with the same logical expression as in step 1 and then using `sum()` to count the successes (again, use `na.rm = TRUE` parameter).
3. Divide the number of successful female applicants by the total number of female applicants.

<div class="notes">

```{r}
n_fem <- sum(appl$gender == 'f', na.rm = TRUE)
n_success_fem <- sum(appl$success[appl$gender == 'f'], na.rm = TRUE)
n_success_fem / n_fem
```
</div>


## Finding out the data type

How can you find out the data type of an object `x`?

- A: `type(x)`
- B: `typeof(x)`
- C: `class(x)`
- D: `help(x)`

<small>Multiple options may apply.</small>

## Finding out the data type

How can you find out the data type of an object `x`?

- A: `type(x)`
- B: `typeof(x)`
- **C: `class(x)` ✓**
- D: `help(x)`

## Finding out the structure of a data frame

How can you find out the structure of a data frame `x`, i.e. an information showing its number of variables and observations and the data type of each variable?

- A: `str(x)`
- B: `help(x)`
- C: `info(x)`
- D: `summary(x)`

<small>Multiple options may apply.</small>

## Finding out the structure of a data frame

How can you find out the structure of a data frame `x`, i.e. an information showing its number of variables and observations and the data type of each variable?

- **A: `str(x)` ✓**
- B: `help(x)`
- C: `info(x)`
- D: `summary(x)`

## You have the following data...

You have the following data in an object `score`:

```{r, echo=FALSE}
(score <- c('3', '1', '1', '2', '3', '5'))
```

You want to know the mean score. Why does `mean(score)` not work? How can you make it work?

## You have the following data...

You have the following data in an object `score`:

```{r, echo=FALSE}
(score <- c('3', '1', '1', '2', '3', '5'))
```

You want to know the mean score. Why does `mean(score)` not work? How can you make it work?

It doesn't work because `score` is not numeric but a character string vector. We have to convert it first:

```{r}
mean(as.numeric(score))    # as.integer() would also be fine here
```

## Getting help

How do you look up the R help page for the function `mean()`?

- A: `google(mean)`
- B: `?mean`
- C: `help(mean)`
- D: `summary(mean)`

<small>Multiple options may apply.</small>

## Getting help

How do you look up the R help page for the function `mean()`?

- A: `google(mean)`
- **B: `?mean` ✓**
- **C: `help(mean)` ✓**
- D: `summary(mean)`

## What's the result...

What's the result of the following statement:

`sum(3:0)`

- A: It's undefined.
- B: 6.
- C: A vector containing 3 and 0.
- D: 0.

<small>Multiple options may apply.</small>

## What's the result...

What's the result of the following statement:

`sum(3:0)`

- A: It's undefined.
- **B: 6. ✓**
- C: A vector containing 3 and 0.
- D: 0.

## Logic

What's the difference between a logical expression (like `x > 5`) and a logical vector (like `c(TRUE, TRUE, FALSE, TRUE)`)?

- A: a logical vector can be the result of a logical expression
- B: there is no difference
- C: I can only subset with a logical vector, not with a logical expression
- D: a logical expression can be the result of a logical vector

<small>Multiple options may apply.</small>

## Logic

What's the difference between a logical expression (like `x > 5`) and a logical vector (like `c(TRUE, TRUE, FALSE, TRUE)`)?

- **A: a logical vector can be the result of a logical expression ✓**
- B: there is no difference
- C: I can only subset with a logical vector, not with a logical expression
- D: a logical expression can be the result of a logical vector

## Indexing

Assume you have the following vector `x`:

```{r}
(x <- letters[1:7])
```

What's the output of the statement `x[-c(2:4)]`?

- A: `"a" "e" "f" "g"`
- B: `"b" "c" "d"`
- C: `"a" "c" "e" "f" "g"`
- D: No output, the statement is invalid R code.

## Indexing

Assume you have the following vector `x`:

```{r}
(x <- letters[1:7])
```

What's the output of the statement `x[-c(2:4)]`?

- **A: `"a" "e" "f" "g"` ✓**
- B: `"b" "c" "d"`
- C: `"a" "c" "e" "f" "g"`
- D: No output, the statement is invalid R code.

## Subsetting data frames {.smaller}

Consider the `airquality` data set:

```{r}
head(airquality, 3)
```

What's the effect of the following statement (assuming that we already loaded the required packages)?

```{r, eval=FALSE}
filter(airquality, Month = 5, Day != 2)
```
<br>

- A: We'll get a subset of the data frame containing only the `Month` and `Day` variables.
- B: We'll get a subset of the data frame only with observations from May, additionally excluding the 2nd of May.
- C: We'll get an error message.
- D: We'll get an empty result data frame.


## Subsetting data frames {.smaller}

Consider the `airquality` data set:

```{r}
head(airquality, 3)
```

What's the effect of the following statement (assuming that we already loaded the required packages)?

```{r, eval=FALSE}
filter(airquality, Month = 5, Day != 2)
```
<br>

- A: We'll get a subset of the data frame containing only the `Month` and `Day` variables.
- B: We'll get a subset of the data frame only with observations from May, additionally excluding the 2nd of May.
- **C: We'll get an error message. ✓**
- D: We'll get an empty result data frame.

